\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms

### 2013-05-16 paper/paper.tex
Added trace collection section

We used a trace-based approach to collect data, and then replayed that data to the rate-selection algorithms using a Python framework.

@@

Our initial plan was to use the algorithms as implemented in the MadWifi drivers for the Atheros chipset. However, MadWifi is deprecated, and only runs on quite old network cards. In addition to the lack of compatible hardware available to us, MadWifi is not an accurate reflection of what a modern user would use. We looked into the ath5k and ath9k projects, which were created to replace MadWifi. Both of these did not port the old bit-rate selection system from MadWifi and instead opted to use the Linux kernel's rate selection software, which only offers Minstrel. Porting SampleRate over to the new drivers would not have been trivial, as there are vast differences in the assumed interfaces.

At this point, we decided to take a different approach and use an approach similar to that used in Sprout~\cite{sprout}.

@@

TODO

\begin{figure}[hbt]
  \hspace{-0.5em}\includegraphics[width=3.5in]{bsmt13floorplan.png}\vspace{-0em}
  \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
\label{figure:1}
\end{figure}

@@

TODO: Harness.py, etc.

We implemented SampleRate as outlined in John Bicket's Master's thesis~\cite{samplerate}. This implementation is slightly different from how it was implemented in the MadWifi kernel driver. The primary difference is that the kernel implementation uses a EWMA instead of a 10 second window. The thesis implementation throws out old rate data and begins anew every 10 seconds. It would be worth looking at the performance of EWMA-based SampleRate in the future.

\texttt{samplerate.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{RemoveStaleResults()}, as well as a few helper functions and data structures for tracking rate statistics.

@@

In SampleRate, the number of successive retries, the frequency of probe packets, the window size, and the lossless $tx\_time$ estimations are all parameters that can be changed. The SampleRate thesis chose 4 successive retries, a probe frequency of every ten packets, and a window size of ten seconds. We used the same values in our implementation. Additionally, we use the same equation to calculate $tx\_time$i.

In Minstrel, the frequency of probe packets, the window size, and the EWMA weighting are the adjustable parameters. The 3.3.8 Linux kernel uses a probe frequency of 10\%, a window size of 100ms, and a EWMA weighing of 0.75. We use the same values. Additionally, Minstrel also estimates $tx\_time$ using the \texttt{ieee80211\_frame\_duration()} method from the \texttt{util.c} file in \texttt{net/mac80211}. We implemented the same function in Python.

@@

Minstrel and SampleRate perform similarly, usually within 3 Mbps of each other. In especially noisy situations, such as when the client is moving or when the client is around the corner from the access point, Minstrel consistently outperforms SampleRate. Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected. This is consistent with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential throughput gains in stable situations. Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}. We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, though, with throughput being up to 67\% higher.

In Figure~\ref{figure:2} we show the performance of constant bit-rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point. The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps. The next highest bit-rate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps. This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}. It demonstrates just how important bit-rate selection is---the optimal bit-rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.

bit-rate selection algorithms strive to have a throughput near or greater than the optimal constant rate. For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate. In noisier situations, though, Minstrel tends to be closer to the optimal constant rate.

We did notice that Minstrel occasionally has very poor performance on traces 1 seconds or shorter---sometimes as low as 2 Mbps. This is because it does not stop sending after four successive failures, but must wait until the end of the 100ms window for statistics to be compiled. Thus it wastes a lot of time sending at non-optimal rates, and a short trace only has a few windows. This behavior does not show up in long traces at all, however, as Minstrel has had time to both smooth out bad windows and collect more statistics (making it less likely to choose a bad rate).

bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly. The variations were small, though. On ten runs on the \texttt{clear\_1.dat} trace, the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum variation of +/- 0.12 Mbps. SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps. The variation of results on other traces was similar.

The histograms of which rates the algorithms decided to send at provided some useful insight into the behavior. Table~\ref{table:2} is a histogram of a 35 second corner trace. In this situation Minstrel outperforms SampleRate. The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were successful at any higher rates. SampleRate does not use these rates except for probe packets, which happens once every ten seconds. Minstrel does not have this measure, however, spends a lot of time sending probe packets at these high rates that happen to never be successful. The only reason low rates have fewer probe packets is because they are second in the retry chain. Despite the fact that SampleRate spends less time probing high rates, it is less successful. This is because Minstrel send the most packets at the three most successful bit-rates. SampleRate, however, spends too much time sampling low bit-rates. Additionally, it spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate spend time at 5.5 Mbps??? We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.

Our examination of the histograms showed us that Minstrel spends a lot of time sampling at high rates. In the specific case we examined, they were all probe packets. But sometimes Minstrel would choose 54 Mbps as the rate with the best throughput, which made little sense. A closer investigation revealed that Minstrel would sometimes luck out and get one success sending a probe at 54 Mbps. The probability of success would be 1, so then 54 Mbps would be chosen as the rate with the best throughput. This is unfortunate, but the real harm came in the next window: nearly all of the packets at 54 Mbps would fail, but Minstrel was still considering 54 Mbps to be the best rate! This is because the EWMA did not account for the fact that the second block having sent hundreds more packets at 54 Mbps than the first. Thus the probability of success would drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly. In essence, the probabilities were not declining fast enough after a large number of failures.

In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix the probability problem.

>>>>

We collected traces about the success rates of each bit rate through a modified Linux wireless driver; Minstrel and SampleRate were then re-implemented in Python and replayed on this collected data.

@@

Our initial plan was to use the algorithms as implemented in the MadWifi drivers for the Atheros chipset. However, MadWifi is deprecated, and only runs on now-ancient network cards. In addition to the unavailability of compatible hardware, MadWifi would not be an accurate reflection of contemporary users. Both ath5k and ath9k, which were created to replace MadWifi, did not port the old bit-rate selection system from MadWifi and instead opted to use the Linux kernel's rate selection framework, which only implements Minstrel.  Porting SampleRate over to the new drivers would not have been trivial, as there are vast differences in the assumed interfaces.

At this point, we decided to take a different approach and use an approach similar to that used in Sprout~\cite{sprout}.  We modified the ath9k driver to try bit rates uniformly and report packet successes and failures, then analyze this data in user-space with re-implementations of SampleRate and Minstrel.  Not only did this lessen the re-implementation cost (since the algorithms were re-implemented in user-space in Python, not in kernel-mode in C), but it also provided reproducibility and allowed us to investigate modifications to SampleRate and to Minstrel.

@@

Traces were collected from an ath9k driver modified to sample bit-rates uniformly (forgoing the usual bit rate selection algorithm) and to report successes and failure from each packet.  Each bit-rate was sampled uniformly in time -- thus, more packets were sent at higher bit-rates, since those bit-rates otherwise deliver packets faster.  For each packet sent, the time to send was recorded, and the bit rate for the next packet was chosen to be the bit-rate that had the smallest total transmissions time.  This guarantees approximately uniform selection of bit-rates.  The wireless card used was an Qualcomm Atheros \texttt{AR9285} Wireless Network Adapter.

Traces were collected at an unused access point in a basement of MIT.  Before initiating trace collection, the wireless card was put in monitor mode for a minute to check that no other users were connected to the access point.  None were, guaranteeing that traces would not include delays from carrier sensing or run into problems due to congestion at the access point.  Traffic was simulated by sending 1500 byte packets to the access point's IP address.  Packets were sent over UDP to avoid the TCP congestion window and retry loop; since no users were using the access point, there was no real risk of deteriorating service for other users.  Figure~\ref{figure:1} shows a floor plan of the access point and basement.

Eighteen total traces were collected, grouped into a number of scenarios.  Each trace was 30 seconds to a minute long, and each scenario was repeated twice.  One scenario placed the wireless card with line of sight to the access point, two meters from it.  Another created solid body interference by having one of the authors stand between the wireless device and the access point.  One placed the device around a corner from the access point, creating some potential for multi-path interference and fading.  In one scenario, we moved the wireless device throughout the trace, pacing back and forth in a 10 meter line in front of the access point.  Finally, one five-minute trace was taken, which mixed the above scenarios, first moving around a corner, then standing still, then moving to line-of-sight with the access point.  A few shorter (3 to 15 second long) traces were collected to investigate the startup behavior of bit-rate selection.

\begin{figure}[hbt]
  \centering
  \includegraphics[width=3.5in]{bsmt13floorplan.png}
  \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
\label{figure:1}
\end{figure}

@@

Once collected, traces were fed to a bit rate selection simulator to analyze the performance of Minstrel and SampleRate.  Both algorithms were represented by two functions: \texttt{ApplyRate}, which returned a multi-rate retry chain of rates to attempt transmission at, and \texttt{ProcessFeedback}, which was passed the success or failure of the transmission and how many attempts were made at each bit rate (along with the current time and the total time to send the packet); this parallels the interface used by the Linux kernel itself.  Helper functions, like SampleRate's \texttt{RemoveStaleResults} function, were instead called from one of \verb|apply_rate| or \verb|process_feedback|.  For every packet transmission, the simulator would call \texttt{ApplyRate}; compute the probability of success based on packets sent in 10-millisecond window around the current time; and from that determine the success or failure of the packet, and the number of retries required.  The results of a test run are thus non-deterministic; both algorithms tested are non-deterministic anyway, so this is no handicap.  To improve re-implementation fidelity, both implementations were meticulously checked by both authors.

We implemented SampleRate as outlined in John Bicket's Master's thesis~\cite{samplerate}. This implementation is slightly different from how it was implemented in the MadWifi kernel driver. The primary difference is that the kernel implementatiion uses a EWMA instead of a 10 second window. The thesis implementation computes success probabilities based on a 10-second sliding window. It would be worth looking at the performance of EWMA-based SampleRate in the future.  Our implementation, \texttt{samplerate.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{RemoveStaleResults()}, as well as a few helper functions and data structures for tracking rate statistics.

@@

In SampleRate, the number of successive retries, the frequency of probe packets, the window size, and the lossless $tx\_time$ estimations are all parameters that can be changed. The SampleRate thesis chose 4 successive retries, a probe frequency of every ten packets, and a window size of ten seconds. We used the same values in our implementation. Additionally, we use the same equation to calculate $tx\_time$.

In Minstrel, the frequency of probe packets, the window size, and the EWMA weighting are the adjustable parameters. The 3.3.8 Linux kernel uses a probe frequency of 10\%, a window size of 100ms, and a EWMA weighting of 0.75. We use the same values. Additionally, Minstrel also estimates $tx\_time$ using the \texttt{ieee80211\_frame\_duration()} method from the \texttt{util.c} file in \texttt{net/mac80211}. We implemented the same function in Python.

@@

Minstrel and SampleRate perform similarly, usually within 3 Mbps of each other. In especially noisy situations, such as when the client is moving or when the client is around the corner from the access point, Minstrel consistently outperforms SampleRate. Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected. This is consistent with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential throughput gains in stable situations. Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}. We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, sometimes with throughput as much as 67\% higher.

In Figure~\ref{figure:2} we show the performance of constant bit-rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point. The optimal constant rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps. The next highest bit-rate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps. This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}. It demonstrates just how important bit-rate selection is---the optimal bit-rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.

We did notice that Minstrel occasionally has very poor performance on traces 1 seconds or shorter---sometimes as low as 2 Mbps. This is because it does not stop sending after four successive failures, but must wait until the end of the 100ms window for statistics to be compiled. Thus it wastes a lot of time sending at non-optimal rates; and a short trace has only a few 100ms windows. This behavior is not evident in long traces, however, as Minstrel has had time to both smooth out bad windows and collect more statistics (making it less likely to choose a bad rate).

Bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly. The variations were small, though. On ten runs on the \texttt{clear\_1.dat} trace, the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum variation of +/- 0.12 Mbps. SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps. The variation of results on other traces was similar.

Additionally, Minstrel uses a complex method of calculating when to send a sample packet that involves a heuristic to make sure that 10\% of packets were actually samples. This is because the mrr places sample packets second in the retry chain when the sample rate is slower than the rate with the best throughput. So even if 10\% of probes have a sample rate in the retry chain, a sample rate isn't actually sent 10\% of the time. We implemented this heuristic and it tended to cause Minstrel to send too many probe packets.

In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix problems with EWMA and too-frequent probing
